---
output: github_document
---

# SPIChanges

<!-- badges: start -->

[![R-CMD-check](https://github.com/gabrielblain/SPIChanges/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/gabrielblain/SPIChanges/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

A package designed to improve the interpretation of the Standardized Precipitation index (SPI) under changing climate conditions.

## Basic Description

The `SPIChanges` was created to detect changes in rainfall patterns and quantify how they affect the probability of a drought event, quantified by the SPI (Mckee et al. 1993), occurring. The package applies a nonstationary approach proposed by Blain et al. (2022), designed to isolate the effect of such changes on the central tendency and dispersion of the SPI frequency distributions.

The package depends on R (\>= 2.10) and imports the following packages: lubridate, zoo, gamlss, gamlss.dist, stats, spsUtil, dplyr, and MuMIn.

## Installation

You can install the development version of SPIChanges from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("gabrielblain/SPIChanges")
```

## Theoretical Background

The SPI was originally designed as a stationary index, where the parameters of the distribution used to estimate the cumulative probability of rainfall quantities do not vary over time. However, due to the alteration of rainfall patterns worldwide from climate change, using nonstationary distributions to calculate a nonstationary version of this drought index (NSPI) has become an important step to adjust for possible changing rainfall patterns (Park et al. 2018). Previous studies, however, have shown that interpreting NSPI estimates is not straightforward, as time-varying parametric distributions remove the effect of rainfall trends on NSPI estimates (Shiau, 2020). Consequently, in the presence of decreasing (or increasing) rainfall trends, the NSPI may underestimate (or overestimate) drought occurrence (Park et al., 2018; Shiau, 2020).

This package, based on the study of Blain et al. (2022), demonstrates that the challenge of interpreting SPI estimates under changing climatic conditions can be overcome by using information generated by the NSPI calculation algorithm. This package uses nonstationary distributions to detect trends in rainfall patterns and quantify their effect on the occurrence of specific SPI values. In other words, the `SPIChanges` package leverages the nonstationary approach of the NSPI algorithm to detect trends in rainfall series and enhance the understanding of how changes in rainfall patterns affect the expected frequency of drought occurrence. With this background, the key function of this package - `SPIChanges()` - is based on the following steps:

1.  Given a rainfall series with data accumulated at a particular time scale, the package uses the stationary version of the two-parameter gamma distribution to calculate SPI estimates and the cumulative probability of each rainfall amount occurring (i.e., the probability of each SPI estimate occurring).

2.  Using the second-order Akaike Information Criterion (AICc), the package selects the best-fitting model among distinct nonstationary two-parameter gamma distributions (see details).

3.  The package uses the selected nonstationary model to calculate the cumulative probability of each rainfall amount occurring under changing climate conditions.

4.  The package compares the probabilities estimated in steps 1 and 3 to indicate whether the frequency of drought or wet events, quantified by the SPI, has increased or decreased over time.

## Basic Instructions

```{r, echo=FALSE, include=FALSE}
# Load the library for examples below (not included in output)
library(SPIChanges)
```

## Function `TSaggreg()`

Aggregates daily rainfall totals at quasiWeek time scales.

## Usage

```{r, eval=FALSE}
TSaggreg(daily.rain,
       start.date,
       TS=4)
```

## Arguments

-   daily.rain: Vector, 1-column matrix or data frame with daily rainfall totals.
-   start.date: Date at which the aggregation should start. Formats: YYYY-MM-DD or YYYY/MM/DD.
-   TS: Time scale on the quart.month basis (integer values between 1 and 96). Default is 4, which corresponds to the monthly time scale.

## Value

Rainfall amounts aggregated at the time scale selected by the user.

## Example 1

Aggregating daily rainfall values at 4-quasiWeek time scale, which corresponds to monthly data.

```{r example 1}
library(SPIChanges)
daily.rain <- CampinasRain[,2]
head(TSaggreg(daily.rain=daily.rain,start.date="1980-01-01",TS=4))
```

## Function `SPIChanges()`

Detect trends and quantify their effect on the probability of SPI values occurring.

## Usage

```{r, eval=FALSE}
SPIChanges(rain.at.TS,
         only.linear)
```

## Arguments

-   rain.at.TS: Vector, 1-column matrix or data frame with rainfall totals accumulated at a time scale.
-   only.linear: A character variable (Yes or No) defining if the function must consider only linear models (Yes) or linear and non-linear models (No). Default is Yes.

## Value

A list object with:

data.week: The precipitation amounts, SPI, cumulative probability of the SPI values (stationary approach), cumulative probability of the precipitation values calculated from the nonstationary algorithm of the NSPI, and the changes in the frequency of the SPI values caused by the changes in precipitation patterns.

model.selection: The generalized additive model (section 2.2) that best fits the precipitation series.

Changes.Freq.Drought: changes in the frequency of moderate, severe and extreme drought events, as definied by the SPI classification system (Table 1), caused by the changes in precipitation patterns.

Statistics: Year to year changes in the expected frequency of moderate, severe and extreme drought events.

## Example 2

Using only linear nonstationary parametric models to assess the probability of rainfall amounts. The models are based on the two-parameter gamma distribution with parameters estimated by the maximum likelihood method. The packages `gamlss` and `gamlss.dist` are used for such estimations.

```{r example 2}
library(SPIChanges)
daily.rain <- CampinasRain[,2]
rainTS4 <- TSaggreg(daily.rain=daily.rain,start.date="1980-01-01",TS=4)
Changes.in.the.SPI <- SPIChanges(rain.at.TS=rainTS4, only.linear = "Yes")
head(Changes.in.the.SPI$data.week)
head(Changes.in.the.SPI$Statistics)
head(Changes.in.the.SPI$model.selection)
head(Changes.in.the.SPI$Changes.Freq.Drought)
```

## Example 3

Using linear and non-linear nonstationary parametric models to assess the probability of rainfall amounts. The models are based on the two-parameter gamma distribution with parameters estimated by the maximum likelihood method. The packages `gamlss` and `gamlss.dist` are used for such estimations.

```{r example 3}
library(SPIChanges)
daily.rain <- CampinasRain[,2]
rainTS4 <- TSaggreg(daily.rain=daily.rain,start.date="1980-01-01",TS=4)
Changes.in.the.SPI <- SPIChanges(rain.at.TS=rainTS4, only.linear = "No")
head(Changes.in.the.SPI$data.week)
head(Changes.in.the.SPI$Statistics)
head(Changes.in.the.SPI$model.selection)
head(Changes.in.the.SPI$Changes.Freq.Drought)
```

# Details

The NSPI uses the same calculation algorithm as the SPI. However, the stationary gamma distribution is replaced by nonstationary versions of this parametric function, which allows its parameters to vary as a function of covariates. In this package, the gamma distributions are fitted to rainfall data using Generalized Additive Models (GAMLSS) with time as a covariate. This fitting process is based on the maximum likelihood method (McCullagh and Nelder, 1989) and considers the following increasingly complex functions (candidate models). Further information on GAMLSS can be found in Rigby and Stasinopoulos (2005).

Model 1 (stationary): The mean (µ) and dispersion (δ) of the distribution are constant over time.

Model 2 (homoscedastic): Only µ is allowed to vary over time linearly.

Model 3: Only δ is allowed to vary over time linearly.

Model 4: Both µ and δ are allowed to vary over time linearly.

Model 5 (homoscedastic): Only µ is allowed to vary over time non-linearly with a quadratic polynomial function.

Model 6: Only δ is allowed to vary over time non-linearly with a quadratic polynomial function.

Model 7: µ is allowed to vary over time non-linearly with a quadratic polynomial function; δ is allowed to vary over time linearly.

Model 8: µ is allowed to vary over time linearly; δ is allowed to vary over time non-linearly with a quadratic polynomial function.

Model 9: Both µ and δ are allowed to vary over time non-linearly with a quadratic polynomial function.

Model 10: Only µ is allowed to vary over time non-linearly with a cubic polynomial function.

Model 11: Only δ is allowed to vary over time non-linearly with a cubic polynomial function.

Model 12: µ is allowed to vary over time non-linearly with a cubic polynomial function; δ is allowed to vary over time linearly.

Model 13: µ is allowed to vary over time linearly; δ is allowed to vary over time non-linearly with a cubic polynomial function.

Model 14: µ is allowed to vary over time non-linearly with a cubic polynomial function; δ is allowed to vary over time non-linearly with a quadratic polynomial function.

Model 15: µ is allowed to vary over time non-linearly with a quadratic polynomial function; δ is allowed to vary over time non-linearly with a cubic polynomial function.

Model 16: µ is allowed to vary over time non-linearly with a cubic polynomial function; δ is allowed to vary over time non-linearly with a cubic polynomial function.

The gamma distribution has two parameters: the shape and scale. Their relationships with the mean and dispersion are given in several studies including Blain et al (2022).

Precipitation frequency distributions are zero-bounded, thus a mixed function that combines the empirical probabilities of no rainfall events (q) and the probability given by the parametric distribution G(x\>0,mu,sigma) must be employed to calculate the SPI (equation 1).\

$$
H(x) = q + (1-q)G(x > 0, \mu, \sigma)      \tag{1}
$$

Within the original SPI algorithm, q is often calculated from equations based on the ratio of zeros in the precipitation records and the sample size. This package uses the binomial distribution to calculate q. As described in studies such as Stagge et al. (2022), this single parameter discrete distribution can also be adapted to nonstationary cases. It was also fitted using GAMLSS models (packages gamlss and gamlss.dist), where zero rain is considered a success and positive rain values are considered failures.

## BugReports:

\<<https://github.com/gabrielblain/SPIChanges/issues> \>

## License:

MIT

## Authors:

Gabriel Constantino Blain, Graciela da Rocha Sobierajski, Leticia Lopes Martins. Maintainer: Gabriel Constantino Blain, [gabriel.blain\@sp.gov.br](mailto:gabriel.blain@sp.gov.br){.email}

## Acknowledgments:

The package uses data from the CPC Global Unified Gauge-Based Analysis of Daily Precipitation data provided by the NOAA PSL, Boulder, Colorado, USA, from their website at https://psl.noaa.gov. The authors greatly appreciate this initiative.

## References

Blain G C, Sobierajski G R, Weight E, Martins L L, Xavier A C F 2022. Improving the interpretation of standardized precipitation index estimates to capture drought characteristics in changing climate conditions. International Journal of Climatology, 42, 5586-5608. <https://doi.org/10.1002/joc.7550>

McCullagh, P. and Nelder, J. A. 1989. Generalized Linear Models. London New York Chapman and Hall. ISBN 9780412317606

Mckee, T. B., Doesken, N.J. and Kleist, J., 1993. The relationship of drought frequency and duration to time scales. In: 8th Conference on Applied Climatology. Boston, MA: American Meteorological Society, 179–184.

Package ‘gamlss', Version 5.4-22, Author Stasinopoulos Mikis et al., <https://CRAN.R-project.org/package=gamlss>

Package ‘gamlss.dist', Version 6.1-1, Author Stasinopoulos Mikis et al., <https://CRAN.R-project.org/package=gamlss.dist>

Park, J., Sung, J.H., Lim, Y-J, Kang, H-S. 2018. Introduction and application of nonstationary standardized precipitation index considering probability distribution function and return period. Theoretical and Applied Climatology, 136:529-542. <https://doi.org/10.1007/s00704-018-2500-y>

Rigby, R.A, Stasinopoulos, D.M. 2005. Generalized additive models for location, scale and shape. Appl Stat 54(3):507–554

Shiau, J-T. 2020. Effects of Gamma-Distribution Variations on SPI-Based Stationary and nonstationary Drought Analyses. Water Resources Management, 34:2081-2095. <https://doi.org/10.1007/s11269-020-02548-x>

Stagge J H and Sung K 2022 A Nonstationary Standardized Precipitation Index (NSPI) using Bayesian splines. Journal of Applied Meteorology and Climatology, 61, 761-779. <https://doi.org/10.1175/JAMC-D-21-0244.1>
