---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# SPIChanges

<!-- badges: start -->
<!-- badges: end -->

A package designed to improve the interpretation of the Standardized Precipitation index under changing climate conditions.

# Basic Description

The package improves the interpretation of the  Standardized Precipitation Index (SPI) under changing climate conditions. It is based on the nonstationary parametric approach proposed in Blain et al. (2022). This latter study developed a four-step algorithm able to detect trends in rainfall patterns and to quantify the effect of such trends on the probability of a SPI value occurring. 

## Installation

You can install the development version of SPIChanges from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("gabrielblain/SPIChanges")
```

## Example 1

Aggregating daily rainfall values at 4-quasiWeek time scale, which corresponds to monthly data.

```{r example 1}
library(SPIChanges)
daily.rain <- CampinasRain[,2]
rainTS4 <- TSaggreg(daily.rain=daily.rain,start.date="1991-01-01",TS=4)
View(rainTS4)
```

## Example 2

Using only linear non-stationary parametric models to assess the probability of rainfall amounts. The models are based on the two-parameter gamma distribution with parameters estimated by the maximum likelihood method. The packages `gamlss` and `gamlss.dist` are used for such estimations.  

```{r example 2}
#
library(SPIChanges)
daily.rain <- CampinasRain[,2]
rainTS4 <- TSaggreg(daily.rain=daily.rain,start.date="1991-01-01",TS=4)
Changes.in.the.SPI <- SPIChanges(rain.at.TS=rainTS4, only.linear = "Yes")
View(Changes.in.the.SPI$data.week)
View(Changes.in.the.SPI$data.week)
View(Changes.in.the.SPI$model.selection)
View(Changes.in.the.SPI$Changes.Freq.Drought)
```
## Example 3

Using linear and non-linear non-stationary parametric models to assess the probability of rainfall amounts. The models are based on the two-parameter gamma distribution with parameters estimated by the maximum likelihood method. The packages `gamlss` and `gamlss.dist` are used for such estimations.  

```{r example 3}
#
library(SPIChanges)
daily.rain <- CampinasRain[,2]
rainTS4 <- TSaggreg(daily.rain=daily.rain,start.date="1991-01-01",TS=4)
Changes.in.the.SPI <- SPIChanges(rain.at.TS=rainTS4, only.linear = "No")
View(Changes.in.the.SPI$data.week)
View(Changes.in.the.SPI$data.week)
View(Changes.in.the.SPI$model.selection)
View(Changes.in.the.SPI$Changes.Freq.Drought)
```

## BugReports:

\<<https://github.com/gabrielblain/SPIChanges/issues> \>

## License:

MIT

## Authors:

Gabriel Constantino Blain, Graciela da Rocha Sobierajski.
Maintainer: Gabriel Constantino Blain, [gabriel.blain\@sp.gov.br](mailto:gabriel.blain@sp.gov.br){.email}

## Acknowledgments:

The package uses data from the Agronomic Institute of Campinas.
The authors greatly appreciate this initiative.

## References

Blain G C, Sobierajski G R, Weight E, Martins L L, Xavier A C F 2022. Improving the interpretation of standardized precipitation index estimates to capture drought characteristics in changing climate conditions. International Journal of Climatology, 42, 5586-5608. <https://doi.org/10.1002/joc.7550>

Package ‘gamlss', Version 5.4-22, Author Stasinopoulos Mikis et al., <https://CRAN.R-project.org/package=gamlss>

Package ‘gamlss.dist', Version 6.1-1, Author Stasinopoulos Mikis et al., <https://CRAN.R-project.org/package=gamlss.dist>
