---
title: "Introduction to SPIChanges"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
   bibliography: bibliography.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The standardized precipitation index (SPI) (@Mckee1993) is a probability-based drought indicator that categorizes dry and wet events as a function of their probability of occurrence or expected frequencies. This widely used drought index was initially developed as a stationary method in which the parameters of the parametric distribution used to estimate the cumulative probability of its unique input variable (rainfall), do not vary over time (@Blain2022). However, there have been observed changes in drought frequency and intensity worldwide (@Strzepek2010, @Dai2012, @Spinoni2019, @Stagge2022, and @Blain2022), which violate the assumption of stationarity (@Coles2001, @Zhang2004, @Cheng2014).

In the context of the current climate changes, studies such as (@Russo2013, @Li2015, and @Rashid2019) proposed the nonstationary standardized precipitation index (NSPI), which incorporates several indices (e.g. time) as external covariates. Therefore, the NSPI was designed to capture nonstationary characteristics of rainfall and, thereby, to capture drought signals under transient rainfall distribution patterns (@Blain2022). However, the use of the NSPI is not a straightforward task (@Blain2022). Because time-varying parametric distributions account for temporal changes in rainfall distributions, they remove the effect of rainfall trends on the NSPI estimates (@Shiau2020). For instance, by applying the NSPI to rainfall series presenting significant increasing trends, (@Shiau2020) noted that the NSPI indicated an increase in the frequency and severity of drought events. Similar results were observed by (@Park2018) that found that, in the presence of decreasing rainfall trends, the use of time-varying distribution leads to an increase in the cumulative probability of a particular rainfall event over time, also increasing the corresponding NSPI estimate. Although this contradictory behavior (@Shiau2020) may be algebraically explained by the NSPI calculation algorithm (@Blain2022), it potentially prevents widespread acceptance of nonstationary distributions for calculating standardized drought indices such as the SPI. This statement particularly holds for the operational use of these indices.

In spite of the above-mentioned limitation, the nonstationary approach adopted by the NSPI algorithm is capable of modelling how the probabilities of rainfall events have changed over a time series. In addition, the use of nonstationary distributions enables isolate the effect of trends on the central tendency and on the dispersion of the rainfall frequency distributions. In this context, the `SPIChanges` package demonstrates that the challenge of interpreting SPI estimates under changing climatic conditions can be overcome by using information generated by the NSPI calculation algorithm. This package uses nonstationary distributions to detect trends in rainfall patterns and quantify their effect on the occurrence of any SPI estimate. In other words, the `SPIChanges` package leverages the nonstationary approach of the NSPI algorithm to detect trends in rainfall series and enhance the understanding of how changes in rainfall patterns affect the expected frequency of drought occurrence. See Theoretical Background for further information on the `SPIChanges` package.

# Getting Started

Load the library in your R session.

```{r setup}
library(SPIChanges)
```

## Data

As described in the introduction, the SPI requires only rainfall data as its input variable. The `SPIChanges` package includes daily rainfall data (`CampinasRain`) to provide use case for its functions. The daily rainfall amounts were measured in millimetres, and recorded in Campinas state of Sao Paulo, Brazil (1980-2023). This data series were obtained from the CPC Global Unified Gauge-Based Analysis of Daily Precipitation, provided by the NOAA/OAR/ESRL PSL, Boulder, CO <https://psl.noaa.gov/>. The data is included in the package as a dataframe with 2 columns and 16071 rows, where the first and second columns are the Date and rainfall amounts, respectively.

## Function `TSaggreg()`

This function aggregates or accumulates daily rainfall totals at quasiWeek time scales.

As a multi-scalar index, the SPI accumulates rainfall totals (often recorded at the daly scale) at any time scale selected by the users. The `SPIChanges` package employs a basic time scale (quasiWeek) that splits each month into four sub-periods. Therefore, a quasiWeek time scale (TS) equal to 4 corresponds to a moving window with a 1-month length. TS equal to 48 corresponds to a moving window with a 1-year length. TS equals to 1 corresponds 1-quasiWeek time scale.\
At this point, it is worth emphasizing that the use of calendar weeks as a basic time scales may pose challenges since the SPI is a relative metric that require homogeneous periods (@Vicente-Serrano2022). The first day of each year can fall on different weekdays, causing inconsistency throughout the year. Leap years may also complicate this comparison in routine drought monitoring (@Vicente-Serrano2022). Additionally, the 4-quasiWeek time scale aligns with the widely used 1-month time scale. 

\## Usage

```{r, eval=FALSE}
TSaggreg(daily.rain,
       start.date,
       TS=4)
```

## Arguments

-   daily.rain: Vector, 1-column matrix or data frame with daily rainfall totals.
-   start.date: Date at which the aggregation should start. Formats: YYYY-MM-DD or YYYY/MM/DD.
-   TS: Time scale on the quart.month basis (integer values between 1 and 96). Default is 4, which corresponds to the monthly time scale.

## Value

Rainfall amounts aggregated at the time scale selected by the user.

## Example 1

Aggregating daily rainfall values at 4-quasiWeek time scale, which corresponds to monthly data.

```{r example 1}
library(SPIChanges)
daily.rain <- CampinasRain[,2]
head(TSaggreg(daily.rain=daily.rain,start.date="1980-01-01",TS=4))
```
