---
title: "Introduction to SPIChanges"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
   bibliography: bibliography.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The standardized precipitation index (SPI) (@Mckee1993) is a probability-based drought indicator that categorizes dry and wet events as a function of their probability of occurrence or expected frequencies. This widely used drought index was initially developed as a stationary method in which the parameters of the parametric distribution used to estimate the cumulative probability of its unique input variable (rainfall), do not vary over time (@Blain2022). However, there have been observed changes in drought frequency and intensity worldwide (@Strzepek2010, @Dai2012, @Spinoni2019, @Stagge2022, and @Blain2022), which violate the assumption of stationarity (@Coles2001, @Zhang2004, @Cheng2014). To address the challenge of using the SPI in the context of the current climate changes, (@Russo2013) and (@Li2015) proposed the non-stationary standardized precipitation index (NSPI), which incorporates several indices (e.g. time) as external covariates (@Rashid2019). Therefore, the NSPI was designed to capture non-stationary characteristics of rainfall and, thereby, to capture drought signals under altered rainfall distribution patterns (@Blain2022). However, the use of NSPI in these types of assessments is not a straightforward task (@Blain2022). Because time-varying parametric distributions account for temporal changes in rainfall distributions, they remove the effect of rainfall trends on the NSPI estimates (@Shiau2020). For instance, by applying the NSPI to rainfall series presenting significant increasing trends, this latter author noted that the NSPI indicated an increase in the frequency and severity of drought events. Similar results were observed by (@Park2019). These authors found that, in the presence of decreasing rainfall trends, the use of time-varying distribution leads to an increase in the cumulative probability of a particular rainfall event over time, also increasing the corresponding NSPI estimate. Although this contradictory behavior (@Shiau2020) may be algebraically explained by the NSPI calculation algorithm, it potentially prevents widespread acceptance of non-stationary distributions for calculating standardized drought indices such as the SPI.

In spite of the above-mentioned limitation, the non-stationary approach adopted by the NSPI algorithm is capable of modelling how the probabilities of rainfall events have changed over a time series. In addition, the use of non-stationary distributions enables isolate the effect of trends on the central tendency and on the dispersion of the rainfall frequency distributions. In this context, the `SPIChanges` package demonstrates that the challenge of interpreting SPI estimates under changing climatic conditions can be overcome by using information generated by the NSPI calculation algorithm. This package uses non-stationary distributions to detect trends in rainfall patterns and quantify their effect on the occurrence of specific SPI values. In other words, the `SPIChanges` package leverages the non-stationary approach of the NSPI algorithm to detect trends in rainfall series and enhance the understanding of how changes in rainfall patterns affect the expected frequency of drought occurrence. In other words, the `SPIChanges` was created to detect changes in rainfall patterns and quantify how they affect the probability of a drought event, quantified by the SPI occurring.

# Getting Started
Load the library in your R session.

```{r setup}
library(SPIChanges)
```
